Negative Selection in the 20th Century

### 2020-05-14 why-negative-selection.Rmd
Add regressions on PCs

Data is taken from UK Biobank. Polygenic scores were  normalized to mean 0,
variance 1. Scores for ADHD, substance use, alcohol, Alzheimer's disease,
autism, bipolarity, BMI, body fat, caffeine, cannabis use, coronary artery
disease, eating disorders, waist-hip ratio, major depressive disorder,
neuroticism, and schizophrenia were reverse coded. High scores are therefore
"positive" in the loose sense that their target phenotypes would be considered
desirable by many people.

@@

* *total siblings*, the number of full siblings of the respondent (including
  himself or herself).
* The number of *children* ever born to/fathered by the respondent.

@@

@@

Within siblings regressions, effect sizes are smaller when controlling for
principal components -- sometimes much smaller, as in the case of height. 
`r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled" effect sizes have 
a smaller absolute value than the corresponding "raw" effect size. The median
proportion between raw and controlled effect sizes is `r median_prop_sibs`.
Among the children regressions, this no longer holds. Effect sizes are barely
affected by controlling for principal components.

```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children. Scores marked * are reverse coded.", fig.height = 7}

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(x = term, y = estimate, shape = reg.type, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_hline + 
      geom_point(size = 2.5, alpha = 0.8) + 
      labs(y = "Effect size", x = "") +
      facet_wrap(vars(dep.var), ncol = 2, scales = "free") +
      coord_flip() +
      scale_colour_manual(values = pval_palette) +
      scale_shape_manual(values = c("circle", "triangle"), 
        guide = guide_legend(title = "")) +
      theme(
        legend.position = "bottom", 
        legend.margin = margin(2,2,2,2)
      )

>>>>

Data is taken from UK Biobank. Polygenic scores were  normalized to mean 0,
variance 1.

@@

* *siblings*, the number of full siblings in the respondent's sib (including
  himself or herself).
* The number of *children* ever born to/fathered by the respondent.

@@

prop_gens_raw <- res_all_wide[["raw_N children"]]/res_all_wide[["raw_N siblings"]]
prop_gens_controlled <- res_all_wide[["controlled_N children"]]/res_all_wide[["controlled_N siblings"]]
median_prop_gens_raw <- median(prop_gens_raw[prop_gens_raw > 0])
median_prop_gens_controlled <- median(prop_gens_controlled[prop_gens_controlled > 0])
```

@@

However, effect sizes are much smaller for *children* than *siblings*
regressions. Among consistently-signed estimates, the median effect size for
children as a proportion of the effect size for  siblings is 
`r median_prop_gens_raw`, or `r median_prop_gens_controlled` with controls.

In siblings regressions, effect sizes are smaller when controlling for
principal components -- sometimes much smaller, as in the case of height. 
`r sum(controlled_smaller_sibs)` out of `r n_pgs` "controlled" effect sizes have 
a smaller absolute value than the corresponding "raw" effect size. The median
proportion between raw and controlled effect sizes is `r median_prop_sibs`.
Among the children regressions, this no longer holds. Effect sizes are barely
affected by controlling for principal components.

```{r plot-basic-regs, fig.cap = "Effects of polygenic scores on number of siblings/children. Scores are plotted on different scales.", fig.height = 7}

n_regs <- as.double(nrow(res_all))
res_all %>% 
      mutate(
        term    = pretty_names(term),
        term    = fct_reorder(term, estimate, order_abs),
        dep.var = fct_inorder(dep.var) 
      ) %>% 
      ggplot(aes(x = term, y = estimate, shape = reg.type, 
        colour = p.value < 0.05/{{n_regs}})) +
      my_hline + 
      geom_point(size = 2.5, alpha = 0.8) + 
      labs(y = "Effect size", x = "") +
      facet_wrap(vars(dep.var), ncol = 2, scales = "free_x") +
      coord_flip() +
      scale_colour_manual(values = pval_palette) +
      scale_shape_manual(values = c("circle", "triangle"), 
        guide = guide_legend(title = "")) +
      theme(
        legend.position = "bottom", 
        legend.margin = margin(2,2,2,2)
      )

To get a further insight into this we regress *siblings* and *children* on
individual principal components themselves. As Figure \@ref(fig:plot-pc-regs) shows,
there are many more large and significant effects  in siblings regressions.

```{r plot-pc-regs, fig.cap = "Effect of principal components on number of siblings/children. Each dot represents one principal component. Points are jittered on the Y axis.", fig.height = 2}

drake::loadd(res_pcs)
n_regs <- as.double(nrow(res_pcs))

res_pcs %>% 
      mutate(
        estimate = abs(estimate),
        dep.var  = fct_relevel(dep.var, "N children")
      ) %>% 
      ggplot(aes(estimate, dep.var, colour = p.value < 0.05/{{n_regs}})) + 
        geom_jitter(height = 0.15, width = 0) + 
        scale_x_log10(labels = scales::label_number(accuracy = 0.0001)) +
        labs(x = "Effect size (log scale)", y = "") + 
        scale_colour_manual(values = pval_palette)
